---
title: "Week 9 Assignment"
author: "Brad Harbans"
date: "4/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(keyring)
library(tidyverse)
library(jsonlite)
library(DT)
```

## Introduction {.tabset}

In this assignment we have been asked to use the [New York Times Developer API](https://developer.nytimes.com/) to obtain information and pull it into a R data frame. 

### Retrieve API Key
I will be storing the API Key in the local keychain of my computer, I will use the `keyring` package to access the key. 
```{r retrieve-api-key}
api.key<-  keyring::key_get("ny_times_api_key_data607")
```

### Using the Top Stories API
For the first example I will use the Top Stories API, to load the top science stories.

```{r get-data-using-top-stories-api}

url <- paste("https://api.nytimes.com/svc/topstories/v2/science.json?api-key=", api.key)

rawDataFromTopStoriesApi <- fromJSON(url,flatten = T)
resultsDataFrameTopStories <- rawDataFromTopStoriesApi$results

resultsDataFrameTopStories %>% 
  select( title,section, short_url, abstract) %>% 
  datatable()

```
### Display Results Graphically

I will now make a bar graph showing the number of articles per section.
```{r ggplot-section}
resultsDataFrameTopStories %>% ggplot(aes(x=section)) +geom_bar()
```

## Conclusion

As can be seen, it is quite useful and simple to access data using an API that returns JSON.
